---
title: "JSC370 MIDTERM REPORT"
author: "Siyi Zhu  |  [GitHub](https://github.com/Sophiaaa12/JSC370---Project)"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---
# Introduction:
This dataset is about "Traveller volumes by port of entry and month" from Government of Canada (https://open.canada.ca/data/en/dataset/1b1c2b92-b388-47d9-87d4-01aee8d3c3e4/resource/22653cdd-d1e2-4c04-9d11-61b5cdd79b4e). I have called API to load the dataset from 2018-01-01 to 2019-05-24 due to the limit. It includes details such as **date, port of entry, region, mode of transportation, and the volume of crossings**. Understanding border traffic is essential for transportation planning, security, and economic forecasting.

Therefore, my research question will be: **How do seasonal and regional factors affect border crossings in Canada from January 2018 to May 2019?** 

Within this question, there are three three hypothesis I can make:<br>
1. **Border crossings follow a seasonal pattern**, peaking during summer months due to increased tourism and declining in winter.<br>
2. **Land border crossings are the most frequent** due to frequent travel between Canada and the U.S.<br>
3. **Certain regions experience significantly higher border crossings** compared to others, reflecting population density and economic activity.<br>


# Methods:
## How and where the data were acquired?
The data is an open data posted on the website of Government of Canada. It enabled API and I acquired it by calling API in RStudio.

## How you cleaned and wrangled the data?
Since the dataset is large, we should ensure we can deal with data better and efficient in the following steps:<br>
- **Converting date columns** to a standardized format (Rename and Change the variable type to the standardized one).<br>
- **Handling missing values**, ensuring no significant data loss.<br>
- **Remove unusal values or non-make sense values**, ensuring the data is correct. <br>
- **Aggregating data** to analyze trends over time, by region, by transportation mode and etc..

## What tools you used for data exploration?
I used the following R packages for data manipulation, visualization, and time series analysis:<br>
- tidyverse, dplyr, and ggplot2 for data wrangling and visualization.<br>
- forecast, timetk for time series analysis.<br>
- kableExtra to generate summary tables for analysis.

This approach ensures a structured and comprehensive analysis of border crossing trends in Canada.<br>


# Preliminary Results:
```{r, include=FALSE}
library(httr)
library(jsonlite)
library(tidyverse)
library(kableExtra)
# Define the CKAN Data API endpoint
api_url = "https://open.canada.ca/data/en/api/3/action/datastore_search"

# Specify the Resource ID of the dataset
resource_id = "22653cdd-d1e2-4c04-9d11-61b5cdd79b4e"

# Initialize an empty dataframe
df_1 = data.frame()

# Set batch size
batch_size = 1000  # Adjust batch size if needed
offset = 0  # Start at record 0

# Fetch data iteratively
repeat {
  response = GET(api_url, query = list(resource_id = resource_id, limit = batch_size, offset = offset))
  data = content(response, "text") %>% fromJSON(flatten = TRUE)
  
  # Extract records and bind to df
  new_data = as.data.frame(data$result$records)
  
  # Stop if no new data is returned
  if (nrow(new_data) == 0) break
  
  df_1 = bind_rows(df_1, new_data)
  
  # Update offset for next batch
  offset = offset + batch_size
}

```
## Summary statistics by transportation mode
From the summary table, we observe that land border crossings have the highest total volume, while rail crossings are the least frequent. Although air travel crossings are fewer in total than land, their maximum daily volume exceeds land crossings, indicating occasional travel spikes. Marine crossings, while lower in total volume, show notable peak activity, likely reflecting seasonal ferry travel. These variations suggest that seasonality and transportation infrastructure play a crucial role in border traffic trends.

```{r echo=FALSE}
# Summary statistics by transportation mode
summary_table <- df_1 %>%
  group_by(Mode) %>%
  summarise(
    Total_Volume = sum(`Sum of Volume`),
    Average_Volume = mean(`Sum of Volume`),
    Max_Volume = max(`Sum of Volume`)
  )

# Display table
kable(summary_table, caption = "Summary of Border Crossings by Transportation Mode") %>%
  kable_styling(full_width = FALSE)

```



## Data Visualization
### Trend of Border Crossings Over Time
The time series plot of total border crossings from 2018 to 2019 exhibits clear seasonal patterns, with periodic fluctuations suggesting weekly or monthly cycles. The overall trend shows a steady increase in border crossings, particularly during mid-2018, followed by a temporary decline. Notably, peaks occur around mid-year, likely due to increased summer travel, while crossings drop during winter months, reflecting seasonal slowdowns. A sharp drop near the end of 2019 may indicate an external event affecting crossings. The fluctuations throughout the period reinforce the seasonal nature of border traffic, likely influenced by holidays, tourism, and economic cycles. These findings support the hypothesis that seasonal and regional factors play a significant role in shaping border crossing trends.

```{r echo=FALSE}
df_1 <- df_1 %>%
  mutate(Date = as.Date(Date))

df_1 <- df_1 %>%
  filter(`Sum of Volume` >= 0)

df_date <- df_1 %>%
  group_by(Date) %>%
  summarise(`Total Volume` = sum(`Sum of Volume`, na.rm = TRUE))

ggplot(df_date, aes(x = Date, y = `Total Volume`)) +
  geom_line(color = "blue") +
  labs(title = "Trend of Border Crossings Over Time (2018-2019)",
       x = "Date",
       y = "Total Volume") +
  theme_minimal()
```


### Trend of Border Crossings by Region
The bar chart illustrates total border crossings by region, showing Southern Ontario and the Pacific Region as the two most active regions and each has over 30 million crossings. The Greater Toronto Area follows. In contrast, the Quebec, Prairie, Northern Ontario, and Atlantic regions experience progressively fewer crossings, with the Atlantic Region having the lowest total volume. The dominance of Southern Ontario and the Pacific Region suggests that major urban centers and key border points significantly influence travel volumes.

```{r echo=FALSE}
df_region <- df_1 %>%
  group_by(Region) %>%
  summarise(`Total Volume` = sum(`Sum of Volume`, na.rm = TRUE))

ggplot(df_region, aes(x = reorder(Region, -`Total Volume`), y = `Total Volume`, fill = Region)) +
  geom_bar(stat = "identity") +
  labs(title = "Trend of Border Crossings by Region",
       x = "Region",
       y = "Total Volume") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none")
```


### Transportation Mode Trends
Land crossings dominate throughout the period, exhibiting strong seasonal fluctuations with noticeable peaks during mid-year. Air travel remains consistent but significantly lower than land crossings, suggesting a more stable flow of air travelers compared to land-based traffic. Marine crossings show moderate seasonal variations, likely influenced by ferry operations, while rail crossings remain minimal throughout. Certain fluctuations align with seasonal trends, suggesting border activity is strongly affected by tourism and seasonal migration patterns. The sharp drop at the end of the period may indicate an external event temporarily disrupting travel across all modes.

```{r echo=FALSE}
df_mode <- df_1 %>%
  group_by(Date, Mode) %>%
  summarise(`Total Volume` = sum(`Sum of Volume`, na.rm = TRUE), .groups = "drop")

ggplot(df_mode, aes(x = Date, y = `Total Volume`, color = Mode)) +
  geom_line() + 
  labs(title = "Transportation Mode Trends",
       x = "Date",
       y = "Total Volume",
       color = "Mode") +
  theme_minimal()
```



# Summary:
## What you found so far from your data in terms of the formulated question?
For my formulated question **How do seasonal and regional factors affect border crossings in Canada from January 2018 to May 2019?**, I found so far about it:<br>
- Border crossings exhibit clear seasonal variations, peaking in warmer months, suggesting a strong link to tourism and seasonal travel.<br>
- Southern Ontario and the Pacific Region consistently report higher volumes of traffic, likely due to economic and population factors.<br>
- Land travel dominates border crossings, while air travel remains stable but represents a much smaller proportion of total crossings.

## Plan of next steps:
To deepen my analysis, I will conduct monthly and seasonal assessments to quantify the seasonal effects on border traffic. To validate key observations, I plan to apply statistical hypothesis testing:<br>
- A t-test will compare weekday and weekend traffic volumes to assess whether weekends experience significantly higher crossings.<br>
- An ANOVA test will be used to determine whether regional and transportation mode differences in border crossing volumes are statistically significant.<br>
- Time series decomposition will be applied to separate trend, seasonal, and residual components, providing a clearer understanding of cyclical and long-term patterns.

Finally, I will synthesize these results into a comprehensive final report, integrating analysis and conclusions to present a well-supported investigation into the impact of seasonal and regional factors on border crossings.